Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-02-09 Manuscript/HI_Model_ms.Rmd
updated manuscript files and added R script for fire emissions graph

**Running title:** Hawaii carbon balance  
**Keywords:** land use, climate change, carbon balance, Hawaii, scenarios, disturbance, ecosystem model

@@

In this study, we consider four unique scenarios exploring a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions. We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands under each of these four scenarios on an annual basis for the period 2010–2100 [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. The four unique scenarios represent all combinations of two land-use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and RCP 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. The goals of this study were to (a) estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, (b) quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and (c) assess the sensitivity of model estimates to the introduction of a CFE on NPP.

@@

Over the last decade, terrestrial ecosystems of the Hawaiian Islands have been a consistent net sink for atmospheric carbon. Net biome productivity (NBP) averaged 0.64 TgC y^-1^ and ranged from 0.46 to 0.88 TgC y^-1^ across all scenarios for the time period 2011-2019.

>>>>

**Running title:** Hawaiʻi carbon balance  
**Keywords:** land use, climate change, carbon balance, Hawaiʻi, scenarios, disturbance, ecosystem model

@@

Here we consider four unique scenarios exploring a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions. We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands under each of these four scenarios on an annual basis for the period 2010–2100 [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. The four unique scenarios represent all combinations of two land-use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

@@

We estimated that terrestrial ecosystems of the Hawaiian Islands have been a consistent net sink for atmospheric carbon over the last decade (figure 4). Net biome productivity (NBP) averaged 0.64 TgC y^-1^ and ranged from 0.46 to 0.88 TgC y^-1^ across all scenarios for the time period 2011-2019. Based on these LUCAS model estimates, Hawaiian terrestrial ecosystems offset approximately 13% of 2015 statewide CO~2~ emissions from energy production and transportation (5.04 TgC), the State of Hawaiʻi's largest source of greenhouse gas emissions [@hawaii_inventory_2019]. Future projections indicate Hawaiian terrestrial ecosystems will continue to be a net sink for atmospheric carbon if global CO~2~ emissions peak around 2040 and then decline (RCP 4.5), and that reducing the intensity of future land use change in Hawaiʻi will further enhance carbon sink strength. If, however, global emissions continue to rise throughout the 21st century (RCP 8.5), our projections indicate Hawaiian ecosystems will transition from a net sink to a net source of CO~2~ to the atmosphere, with high intensity land-use change accelerating this transition and exacerbating net carbon loss.

Our NBP estimates of ~ 0.64 TgC y^-1^ for the recent past (2011-2019) agree well with those from a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019].
